Assessing Population Density and Crime Patterns in Toronto Neighborhoods (2014–2023)

### 2024-09-27 paper/paper.qmd
Revise some typos.

---
title: "Assessing Population Density and Crime Patterns in Toronto Neighborhoods (2014–2023)"
author: 
  - Yongqi Liu
thanks: "Code and data are available at:https://github.com/Cassieliu77/Crime-Patterns-and-Population-Density.git"
date: 27 September 2024
date-format: long
toc: true 
abstract: "This paper examines the relationship between population density and crime distribution across Toronto neighborhoods from 2014 to 2023. The analysis reveals that high-density areas experience higher crime rates, particularly for auto theft and assault, with socioeconomic factors further influencing crime patterns. Based on these findings, we propose targeted interventions and offers data-driven suggestions for policymakers and urban planners to enhance crime prevention strategies in Toronto."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

# Data {#sec-data}
## Data Overview
The Neighborhood Crime Rates dataset, obtained from @opendatatoronto, spans from 2014 to 2023 and covers 158 Toronto neighborhoods.It is designed to offer communities more information about public safety and awareness. The data provided to the Toronto Police Service by reporting parties is preliminary and may not have undergone full verification at the time of publication. The dataset includes various crime types such as Assault, Auto Theft, Robbery, Break and Enter, Bike Theft, Shooting, and Homicide. According to the standard definition by Statistics Canada, the crime rate is calculated as the number of crimes per 100,000 people each year. Besides, crime counts per crime type per year are also shown in the data. In @opendatatoronto, another variable of interest *Population 2023* refers to the population in the 2023 year in each neighborhood, enabling the analysis of the relationship between population density and crime frequency. Geometry column is transformed into spatial data to build up the Toronto city map to observe population density and crime distribution. @tbl-rawdata gives us a glimpse of how the dataset looks. Each neighborhood is identified by its *Area Name* and a unique *Hood ID*.

The average crime rates represent the mean crime rate across 158 neighborhoods from 2014 to 2023, and were calculated to analyze long-term trends and highlight crime types that have seen significant increases. The variables *Crime Type* and *Crime Count* are based on 2023 data (@tbl-summary1).Each neighborhood’s average rank was determined by ranking all 158 neighborhoods based on their total annual crime count. A rank is assigned for each year, and the average rank over the past decade (2014–2023) is calculated by averaging these yearly ranks. A lower rank indicates a higher crime count. Summaries of the Top 10 and Bottom 10 ranked neighborhoods are provided in @tbl-summary2 and @tbl-summary4.

@@

Crime rates offer a more balanced comparison over time as they account for population changes in the area and give us a fairer way to see the crime trends in the past decade, which explains why a similar dataset called "Police Annual Statistical Report - Reported Crimes" is not be chosen as my dataset, which only contains the reported crime count for a certain year and a certain division. It lack a standard way to show the crime trends and spatial attribute to see the geography distribution and it also includes reported crimes deemed as unfounded.

@@

@fig-rates reveals a significant upward trend in **Auto Theft** and **Assault**, particularly after 2020. The crime rate for **Assault** consistently remained the highest across all crime types, starting at around 600 per 100,000 population in 2014 and reaching a peak of approximately 800 per 100,000 population in 2023. This consistent rise suggests that assault has been a persistent issue in Toronto, with an even sharper increase following the onset of the COVID-19 pandemic. **Break and Enter**,**Robbery** and **Theft Over** all experienced increases after 2021. This rise suggests a sharp increase in private property related crimes, possibly influenced by socioeconomic factors or changes in law enforcement or technology. In contrast, **Homicide** and **Shooting** had relatively lower crime rates and displayed more stability over time, with little significant change compared to **Assault** and **Auto Theft**.

@@

## Toronto Population Density
@fig-Population shows the population density in Toronto, constructed by using variable *Population 2023* and geometry data. Higher saturation represents higher population density in 2023. Neighborhoods with lighter shading have smaller populations, such as neighborhood 12 and 13. Notably, neighborhoods 1 and 95 exhibit the darkest saturation, indicating the highest population density. In the following sections, we will explore whether these high-density neighborhoods correspond to areas that are more prone to crime.

@@

From @fig-crimeandpopulation, most of these scatter plots represent upward trend between Crime Count and Population is shown. It seems that the lack of cases in **Shooting** and **Homicide** makes most dots overlap together. These crimes are less frequent, and the relationship between population and crime count is less pronounced compared to property crimes.

Consequently, extra maps are shown in @fig-homicide and @fig-shooting to display these two relatively rare crime types' occurrence in particular neighborhoods. The size of red dot represents for the crime count and is marked on the area they occur. It can be observed that most larger size dots appear in the neighborhoods with darker shading, and most of neighborhoods with lowest saturation blue does not have a reported homicide or shooting case.

@@

## Neighborhood Analysis
After looking at the pattern of crimes in the whole Toronto, it is more important to take a look into those small neighborhoods. The rank here is counted based on the total crime count for each neighborhood each year. Higher rank implies higher crime count on average. Top 10 safest and Top 10 Crime-Prone neighborhoods are drawn from the average rank in past 10 years. 
Neighborhoods like West Humber-Clairville and York University Heights ranked among the most crime-prone areas, likely due to their high population density (around 30,000 population) and busy commercial districts. This supports existing criminological theories that link high-density environments with increased opportunities for crime. Besides, as we noted in @fig-Population, both of the most density neighborhoods 1 (West Humber-Clairville) and 95 (Annex) show in the Top 10 crime-prone ranked neighborhoods. In addition to West Humber-Clairville and Annex, Wellington Place, Kensington-Chinatown, and Moss Park also ranked among these most crime-prone areas over the past decade. These neighborhoods are characterized by higher population densities and socioeconomic challenges, making them more vulnerable to crime. The increasing trend in auto theft and assault in these neighborhoods underscores the need for targeted interventions and enhanced security measures.

@@

In contrast, @tbl-summary4 shows the neighborhoods with least crime occurrences in past 10 years. Neighborhoods such as Maple Leaf and Lambton Baby Point have consistently ranked as some of the safest areas in Toronto. These neighborhoods tend to have lower population densities, and all of them just have a population around 10,000 in 2023. This significant difference is likely due to their suburban residential character, lower population densities, and enhanced social cohesion.”

@@

##  Targeted Crime Prevention Strategies
A pattern has been verified in Toronto is that persistently high crime rates often draw additional criminal activity, leading to what is known as the “law of crime concentrations at places” [@mohammadi]. It means crimes tend to recur in some areas with specific features, thus these features can be found out to take crime prevention strategies. Property and violent crime hotspots are concentrated in the downtown core, northwest, and eastern parts of the city [@Wang]. This geographic concentration has significant implications for law enforcement and public policy. Policymakers and law enforcement agencies must recognize that crime prevention strategies cannot be one-size-fits-all. Instead, interventions need to be tailored to address the unique challenges faced by high-risk neighborhoods.

This study underscores the importance of community-based crime prevention strategies. Engaging local residents through neighborhood watch programs, educating the public on crime prevention techniques, and fostering collaboration between the police and community organizations can empower communities to take charge of their safety. Such an approach not only helps reduce crime frequency but also strengthens the relationship between law enforcement and the communities they serve—an essential element for long-term success in crime reduction.

@@

##  Limitations and Future Research
While this study provides a strong foundation for assessing population density and crime patterns in Toronto neighborhoods, some limitations must be acknowledged. Firstly, the analysis relies on reported crime data, which may not fully reflect the actual scope of criminal activity, especially for underreported crimes like domestic violence and cybercrime. Information era has led to some new types of criminal activities like cybercrime and telecom crime. Obtaining a definitive national cybercrime rate or count is still a challenging task because of underreporting.

@@

Future research could expand on these findings by incorporating additional data sources and other potential influences to capture a wider range of criminal activity. It would also be valuable to explore how different aspects of urban infrastructure and development of new technology affect crime patterns.

>>>>

---
title: "Assessing Population Density and Crime Patterns in Toronto Neighborhoods (2014–2023)"
author: 
  - Yongqi Liu
thanks: "Code and data are available at:https://github.com/Cassieliu77/Crime-Patterns-and-Population-Density.git"
date: 27 September 2024
date-format: long
toc: true 
abstract: "This paper examines the relationship between population density and crime distribution across Toronto neighborhoods from 2014 to 2023. The analysis reveals that high-density areas experience higher crime rates, particularly for auto theft and assault, with socioeconomic factors further influencing crime patterns. Based on these findings, we propose targeted interventions and offer data-driven suggestions for policymakers and urban planners to enhance crime prevention strategies in Toronto."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

# Data {#sec-data}
## Data Overview
The Neighborhood Crime Rates dataset, obtained from @opendatatoronto, spans from 2014 to 2023 and covers 158 Toronto neighborhoods. It is designed to offer communities more information about public safety and awareness. The data provided to the Toronto Police Service by reporting parties is preliminary and may not have undergone full verification at the time of publication. The dataset includes various crime types such as Assault, Auto Theft, Robbery, Break and Enter, Bike Theft, Shooting, and Homicide. According to the standard definition by Statistics Canada, the crime rate is calculated as the number of crimes per 100,000 people each year. Besides, crime counts per crime type per year are also shown in the data. In @opendatatoronto, another variable of interest *Population 2023* refers to the population in the 2023 year in each neighborhood, enabling the analysis of the relationship between population density and crime frequency. The geometry column is transformed into spatial data to build up the Toronto city map to observe population density and crime distribution. @tbl-rawdata gives us a glimpse of how the dataset looks. Each neighborhood is identified by its *Area Name* and a unique *Hood ID*.

The average crime rates represent the mean crime rate across 158 neighborhoods from 2014 to 2023 and were calculated to analyze long-term trends and highlight crime types that have seen significant increases. The variables *Crime Type* and *Crime Count* are based on 2023 data (@tbl-summary1). Each neighborhood’s average rank was determined by ranking all 158 neighborhoods based on their total annual crime count. A rank is assigned for each year, and the average rank over the past decade (2014–2023) is calculated by averaging these yearly ranks. A lower rank indicates a higher crime count. Summaries of the Top 10 and Bottom 10 ranked neighborhoods are provided in @tbl-summary2 and @tbl-summary4.

@@

Crime rates offer a more balanced comparison over time as they account for population changes in the area and give us a fairer way to see the crime trends in the past decade, which explains why a similar dataset called "Police Annual Statistical Report - Reported Crimes" is not be chosen as my dataset, which only contains the reported crime count for a certain year and a certain division. It lacks a standard way to show the crime trends and spatial attributes to see the geography distribution and it also includes reported crimes deemed as unfounded.

@@

@fig-rates reveals a significant upward trend in **Auto Theft** and **Assault**, particularly after 2020. The crime rate for **Assault** consistently remained the highest across all crime types, starting at around 600 per 100,000 population in 2014 and reaching a peak of approximately 800 per 100,000 population in 2023. This consistent rise suggests that assault has been a persistent issue in Toronto, with an even sharper increase following the onset of the COVID-19 pandemic. **Break and Enter**,**Robbery** and **Theft Over** all experienced increases after 2021. This rise suggests a sharp increase in private property-related crimes, possibly influenced by socioeconomic factors or changes in law enforcement or technology. In contrast, **Homicide** and **Shooting** had relatively lower crime rates and displayed more stability over time, with little significant change compared to **Assault** and **Auto Theft**.

@@

## Toronto Population Density
@fig-Population shows the population density in Toronto, constructed by using variable *Population 2023* and geometry data. Higher saturation represents higher population density in 2023. Neighborhoods with lighter shading have smaller populations, such as neighborhoods 12 and 13. Notably, neighborhoods 1 and 95 exhibit the darkest saturation, indicating the highest population density. In the following sections, we will explore whether these high-density neighborhoods correspond to areas that are more prone to crime.

@@

From @fig-crimeandpopulation, most of these scatter plots represent an upward trend between Crime Count and Population is shown. It seems that the lack of cases in **Shooting** and **Homicide** makes most dots overlap together. These crimes are less frequent, and the relationship between population and crime count is less pronounced compared to property crimes.

Consequently, extra maps are shown in @fig-homicide and @fig-shooting to display these two relatively rare crime types' occurrence in particular neighborhoods. The size of the red dot represents the crime count and is marked on the area they occur. It can be observed that most larger-sized dots appear in the neighborhoods with darker shading, and most of the neighborhoods with the lowest saturation blue do not have a reported homicide or shooting case.

@@

## Neighborhood Analysis
After looking at the pattern of crimes in the whole of Toronto, it is more important to take a look into those small neighborhoods. The rank here is counted based on the total crime count for each neighborhood each year. A higher rank implies a higher crime count on average. The Top 10 safest and Top 10 Crime-Prone neighborhoods are drawn from the average rank in the past 10 years. 
Neighborhoods like West Humber-Clairville and York University Heights ranked among the most crime-prone areas, likely due to their high population density (around 30,000 population) and busy commercial districts. This supports existing criminological theories that link high-density environments with increased opportunities for crime. Besides, as we noted in @fig-Population, both of the density neighborhoods 1 (West Humber-Clairville) and 95 (Annex) show in the Top 10 crime-prone ranked neighborhoods. In addition to West Humber-Clairville and Annex, Wellington Place, Kensington-Chinatown, and Moss Park also ranked among these most crime-prone areas over the past decade. These neighborhoods are characterized by higher population densities and socioeconomic challenges, making them more vulnerable to crime. The increasing trend in auto theft and assault in these neighborhoods underscores the need for targeted interventions and enhanced security measures.

@@

In contrast, @tbl-summary4 shows the neighborhoods with the least crime occurrences in the past 10 years. Neighborhoods such as Maple Leaf and Lambton Baby Point have consistently ranked as some of the safest areas in Toronto. These neighborhoods tend to have lower population densities, and all of them just have a population of around 10,000 in 2023. This significant difference is likely due to their suburban residential character, lower population densities, and enhanced social cohesion.

@@

##  Targeted Crime Prevention Strategies
A pattern that has been verified in Toronto is that persistently high crime rates often draw additional criminal activity, leading to what is known as the “law of crime concentrations at places” [@mohammadi]. It means crimes tend to recur in some areas with specific features, thus these features can be found to take crime prevention strategies. Property and violent crime hotspots are concentrated in the downtown core, northwest, and eastern parts of the city [@Wang]. This geographic concentration has significant implications for law enforcement and public policy. Policymakers and law enforcement agencies must recognize that crime prevention strategies cannot be one-size-fits-all. Instead, interventions need to be tailored to address the unique challenges faced by high-risk neighborhoods.

This study underscores the importance of community-based crime prevention strategies. Engaging residents through neighborhood watch programs, educating the public on crime prevention techniques, and fostering collaboration between the police and community organizations can empower communities to take charge of their safety. Such an approach not only helps reduce crime frequency but also strengthens the relationship between law enforcement and the communities they serve—an essential element for long-term success in crime reduction.

@@

##  Limitations and Future Research
While this study provides a strong foundation for assessing population density and crime patterns in Toronto neighborhoods, some limitations must be acknowledged. Firstly, the analysis relies on reported crime data, which may not fully reflect the actual scope of criminal activity, especially for underreported crimes like domestic violence and cybercrime. The information era has led to some new types of criminal activities like cybercrime and telecom crime. Obtaining a definitive national cybercrime rate or count is still a challenging task because of underreporting.

@@

Future research could expand on these findings by incorporating additional data sources and other potential influences to capture a wider range of criminal activity. It would also be valuable to explore how different aspects of urban infrastructure and the development of new criminal means affect crime patterns.
